---
title: "Predicting Down syndrome from protein expression in mice"
author: "Dayoung Yu (dry2115), Justin Hsie (2119), Christian Pascual (cbp2128)"
date: "4/27/2019"
output: pdf_document
---

# Introduction

Down syndrome is a genetic disorder that originates from a full or partial extra copy of chromosome 21. The hallmarks of this syndrome include intellectual disability, developmental delays and distinct physical characteristics. Currently, there is no cure to Down syndrome, but screening and diagnostic tests exist for the condition. Some diagnostic tests such as amniocentisis have been demonstrated to increase the risk of miscarriage, so prenatal screening is preferred. A systematic review by Yao et. al demonstrated that two prenatal screening tests in serum have sensitivities ranging from 77% to 93%. However, amniocentisis is much more predictive of Down syndrome.

This disparity offers an interesting classification problem: given expression data, can we predict Down syndrome with comparable accuracy to screening and diagnostic tests? We'll use the *Mice Protein Expression* data set from Kaggle to explore this question. In its raw form, the data set contains expression levels for 77 proteins in 1080 mice. These mice were either controls or trisomic, so we can use this data set for classification purposes.

## Data Cleaning

```{r, warning = FALSE, message = FALSE }
library(tidyverse)
library(visdat)
mice = read_csv('./data.csv') %>% 
  mutate(MouseID = 1:length(MouseID))
```

A quick check of the data reveals that it contains missing values. Below we'll visualize how the missing data is distributed.

```{r, fig.width = 5, fig.height = 3.3, dpi = 300, fig.align = 'center'}
vis_miss(mice) + theme(axis.text = element_text(size = 3))
```

Two key observations come from this visualization of the missing data: 1) There is a small subset of mice missing almost half of their data, and 2) there is a small subset of proteins that are missing in most mice. To pare down the data set, we'll remove the mice that are missing most of their proteins and the proteins that are missing in most of the mice. 

```{r}
final.mice = mice %>% 
  filter(!(MouseID %in% c(988, 989, 990))) %>% 
  select_if(function(col) !any(is.na(col))) %>% 
  select(DYRK1A_N:Genotype) %>% 
  mutate(Genotype = ifelse(Genotype == "Control", 0, 1))
```

The resulting data set contains 68 proteins with complete information on 1077 mice. Each of the candidate proteins are on a similar magnitude, so we opted to not center and scale the data. Trisomic status was relabeled numerically to `0|1`.

# References

1. Yao Y, Liao Y, Han M, Li SL, Luo J, Zhang B. Two kinds of common prenatal screening tests for Down's syndrome: a systematic review and meta-analysis. Sci Rep. 2016;6:18866. Published 2016 Jan 6. doi:10.1038/srep18866
